\input{preamble}

\title{Representations of abelian groups}
\author{Arpon Raksit}
\date{December 30, 2015}

\externalgrain{abelian-category-linear-algebra}
\externalgrain{representations-semisimplicity}

\begin{document}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Let $k$ be an algebraically closed field. Let $\cA$ be an abelian $k$-linear category. Let $G$ be an abelian group and let $\cA^G$ denote the category of $G$-representations in $\cA$, i.e. the functor category $[\rB G, \cA]$ (which is also obviously abelian and $k$-linear in a natural way).

\begin{proposition}
  \label{simple-impl-char}
  Let $M$ be a simple object in $\cA^G$. Then there's a character $\lambda \c G \to k^\times$ such that the action of $G$ on $M$ is given by multiplication by $\lambda$.
\end{proposition}

\begin{proof}
  The claim is simply that the action $\gamma(g) \in \End_\cA(M)$ of each $g \in G$ is a multiple $\lambda(g) \in k$ of the identity. Since $G$ is abelian we have
  \[
    \gamma(g)\gamma(h) = \gamma(gh) =
    \gamma(hg) = \gamma(h)\gamma(g)
  \]
  for all $h \in H$. I.e. $\gamma(g)$ is also a morphism of representations. The claim then follows from the formulation of Schur's lemma \cref{abelian-category-linear-algebra:schur-alg-closed} stating that $M$ being simple implies its only endomorphisms are multiplication by scalars.
\end{proof}

\begin{remark}
  \label{simple-impl-simple}
  It follows immediately from \cref{simple-impl-char} that if $M$ is simple in $\cA^G$ then $M$ is also simple when viewed as an object in $\cA$.
\end{remark}

\begin{corollary}
  \label{isotypic}
  Assume $\cA$ is semisimple. Then any representation $M \in \cA^G$ has an \emph{isotypic decomposition}
  \[
    M \iso \bigoplus_{\lambda \in \dual G} M_\lambda,
  \]
  where for $\lambda \in \dual G$ the representation $M_\lambda \in \cA^G$ is acted on by $G$ via $\lambda$, i.e. for $g \in G$ the action $g \c M_\lambda \to M_\lambda$ is given by multiplication by $\lambda(g)$.
\end{corollary}

\begin{proof}
  Since $\cA$ is semisimple, by \cref{representations-semisimplicity:semisimple} we know that $\cA^G$ is semisimple. Thus any $M \in \cA^G$ is a direct sum of simple objects. The claim now follows from \cref{simple-impl-char}.
\end{proof}

\begin{remark}
  \label{isotypic-unique}
  Note that the isotypic decomposition provided by \cref{isotypic} is automatically unique: for $\lambda \in \dual G$ we can recover $M_\lambda$ as the kernel of the map $M \to \bigoplus_{g \in G} M$ which on the factor indexed by $g \in G$ is given by multiplication by $1 - \lambda(g)$.
\end{remark}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \bibliographystyle{amsalpha}
% \bibliography{refs}

\end{document}
