\input{preamble}

\title{Semisimplicity of representations}
\author{Arpon Raksit}
\date{December 30, 2015}

\externalgrain{abelian-category-linear-algebra}

\begin{document}
\maketitle

\newcommand{\chr}{\operatorname{char}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Let $k$ be a field. Let $\cA$ be an abelian $k$-linear category. Let $G$ be a finite group such that $\chr(k)$ does not divide $|G|$. Denote by $\cA^G$ the (abelian $k$-linear) category of $G$-representations in $\cA$, i.e. the functor category $[\rB G, \cA]$.

\begin{lemma}[Averaging trick]
  \label{avg-trick}
  Suppose all short exact sequences in $\cA$ split. Then the same is true in $\cA^G$.
\end{lemma}

\begin{proof}
  Consider a short exact sequence
  \[
    0 \to X' \to X \to X'' \to 0,
  \]
  in $\cA^G$. This is a short exact sequence when viewed in $\cA$ as well, so by hypothesis there is a splitting $T \c X'' \to X$ in $\cA$. Then the average
  \[
    \frac{1}{|G|} \sum_{g \in G} gTg^{-1}
  \]
  is a morphism $X'' \to X$ in $\cA^G$ which also splits the sequence. (Note that this average makes sense by our hypothesis that $\chr(k)$ does not divide $|G|$.)
\end{proof}

\begin{proposition}
  \label{semisimple}
  Suppose $\cA$ is semisimple. Then $\cA^G$ is semisimple.
\end{proposition}

\begin{proof}
  Recall \cref{abelian-category-linear-algebra:semisimple-iff-split} that $\cA$ being semisimple is equivalent to all short exact sequences in $\cA$ splitting and all objects in $\cA$ having finite length. For an object in $\cA^G$ we'll refer to its length in $\cA$ as its \emph{$\cA$-length}.

  Now, we wish to show any nonzero $X \in \cA^G$ is semisimple. We induct on its $\cA$-length. We may choose a nonzero subobject $\phi \c X' \inj X$ in $\cA^G$ of  minimal $\cA$-length. Clearly $X'$ must be simple in $\cA^G$. If $\phi$ is an isomorphism then we're done. Otherwise, consider the short exact sequence
  \[
    0 \to X' \to X \to X'' \to 0
  \]
  in $\cA^G$. By \cref{avg-trick} it splits. Since $X'$ is simple this reduces the semisimplicity of $X$ to the semisimplicity of $X''$. But $X''$ has strictly smaller $\cA$-length than $X$ so we're done by induction.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \bibliographystyle{amsalpha}
% \bibliography{refs}

\end{document}
